---
title: 'SOM: Diagnostics of linear models'
author:
- Ben Marwick
- Elspeth Hayes
- Chris Clarkson
- Richard Fullagar
date: '`r Sys.Date()`'
output:
   bookdown::word_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: trampling.bib
abstract: |
  This is the abstract.
---

```{r, echo = FALSE}
# extract code from main manuscript file into a new R file
# so we can run select chunks from that in this file
library(knitr)
library(ggfortify)
opts_chunk$set(echo = FALSE, 
               messages = FALSE, 
               warnings = FALSE, 
               dpi = 300,
               fig.show = 'hide')
invisible(purl("Marwick_Hayes_et_al.Rmd",  
               output="temp",
               quiet=TRUE))
read_chunk("temp")
```


```{r, ref.label="load_libraries"}
```

```{r, ref.label="load_data"}
```

```{r, ref.label="split_by_trampling_event"}
```

```{r, ref.label="prepare_point_data"}
```

```{r, ref.label="zinggplotofartefacts"}
```

```{r, ref.label = "hordispldists"} 
```

```{r, ref.label = "hordirectionshist"}
```

```{r, ref.label = "hormodels"}
```

## Diagnostic plots for linear models of horizontal displacement of artefacts

```{r, ref.label = "hormodelsdiagnostics"}
```

```{r,  fig.show = 'as.is'}
# draw diagnostic plots
hormodels_diagnostics_plots <- 
 map(hormodels_diagnostics,
     ~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

# view
hormodels_diagnostics_plots


kable(hormodels_diagnostics_influence, 
      caption = "Horizontal displacement: Correlation of artefact volume with Cook's distance values for each model")
  
```

## Diagnostic plots for linear models of horizontal direction of artefacts

```{r, ref.label = "hordirmodelsdiagnostics"}
```

```{r,  fig.show = 'as.is'}

 # draw diagnostic plots
  
hordirmodels_diagnostics_plots <- 
 map(hordirmodels_diagnostics, ~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

# view
hordirmodels_diagnostics_plots



kable(hordirmodels_diagnostics_influence, 
      caption = "Horizontal direction: Correlation of artefact volume with Cook's distance values for each model")
 
```

## Diagnostic plots for linear models of vertical displacement of artefacts


```{r, ref.label = "verdisp"}
```

```{r, ref.label = "vermodels"}
```

```{r, ref.label = "vermodelsdiagnostics"}
```

```{r,  fig.show = 'as.is'}

# draw diagnostic plots
vermodels_diagnostics_plots <- 
  map(vermodels_diagnostics, 
      ~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

vermodels_diagnostics_plots


kable(vermodels_diagnostics_influence, 
      caption = "Vertical displacement: Correlation of artefact volume with Cook's distance values for each model")
```

## Diagnostic plots for linear models of artefact orientation

```{r, ref.label = "verangles"}
```

```{r, ref.label = "artefactorientation"}
```

```{r, ref.label = "artefactplunge"}
```

```{r, ref.label = "orientplungcorr"}
```

```{r, ref.label = "orientmodelsdiagnostics"}
```

```{r,  fig.show = 'as.is'}
orientmodels_diagnostics_plots <- 
  orientmodels_diagnostics %>% 
  map(~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

orientmodels_diagnostics_plots


kable(orientmodels_diagnostics_influence, 
      caption = "Artefact orientation: Correlation of artefact volume with Cook's distance values for each model")
 
```

## Diagnostic plots for linear models of artefact plunge


```{r, ref.label = "plungemodelsdiagnostics"}
```


```{r,  fig.show = 'as.is'}

plungemodels_diagnostics_plots <- 
  plungemodels_diagnostics %>% 
  map(~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

plungemodels_diagnostics_plots


kable(plungemodels_diagnostics_influence, 
      caption = "Artefact plunge: Correlation of artefact volume with Cook's distance values for each model")
 
```

```{r}
# delete the temp file we made at the top of this file
unlink("temp")
```

## Simulation of many trampling events

How to determine the how deep is too deep for trampling to have an effect on an artefact? Here we look at the distribution of distances below the starting point, and assess which theoretical distribution is the best fit. 

```{r}
ggplot(ver.s, aes(displ)) + 
  geom_histogram()

# take the -ve values and fit a distribution
# focus on the distrubution of moves when an artefact moves down
negative_vert <- ver.s$displ[ver.s$displ < 0]

library(fitdistrplus)
library(logspline)

descdist(negative_vert, discrete = FALSE)

possible_distributions <- c("lnorm", "gamma",  "beta", "exp", "norm")

# compute fits to distributions
fits <- map(possible_distributions, ~fitdist(-negative_vert, .x))

# for visual inspections
map(fits, ~plot(.x))

# find model with lowest AIC
best_distribution <- 
map_dbl(fits, ~.x$aic) %>% 
  which.min() %>% 
  possible_distributions[.]

best_distribution_details <- 
fits[[which(possible_distributions == best_distribution)]]
```

We evaluated `r length(possible_distributions)` possible distributions to find the best fit with our observed values of artefact displacement below the point of origin. Using visual inspection of diagnostic plots and the AIC valuess, we have determined that a `r best_distribution` distribution is the best fit for the observed vertical displacenent values.


```{r}
# generate a bunch of fitted gamma distributions
n <- 1000
many_fitted_distributions <- 
replicate(n, rgamma(n = length(negative_vert), 
                      shape = best_distribution_details$estimate[[1]], 
                      rate= best_distribution_details$estimate[[2]]))



data_to_plot <- 
   data.frame(many_fitted_distributions,
            observed_data = -negative_vert) %>% 
   gather(variable, 
          value)

ggplot() +
   geom_density(data = filter(data_to_plot, 
                              grepl("X", variable)),
                aes(value, 
                    group = variable),
               colour = alpha("grey70", 0.1)) +
  geom_density(data = filter(data_to_plot, 
                             variable == 'observed_data'),
               aes(value),
               colour = alpha("red", 1)) +
   theme_minimal()
```

The plot above shows `r n` simulated `r best_distribution` distributions (in grey) and our observed distrubtion in red. 

```{r}
ninety_nine_point_five <- 
qgamma(0.995, 
       shape = best_distribution_details$estimate[[1]],
       rate = best_distribution_details$estimate[[2]])
```

