---
title: 'SOM: Diagnostics of linear models'
author:
- Ben Marwick
- Elspeth Hayes
- Chris Clarkson
- Richard Fullagar
date: '`r Sys.Date()`'
output:
   bookdown::word_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: trampling.bib

---

```{r, echo = FALSE}
# extract code from main manuscript file into a new R file
# so we can run select chunks from that in this file
library(knitr)
library(ggfortify)
opts_chunk$set(echo = FALSE, 
               messages = FALSE, 
               warnings = FALSE, 
               dpi = 300,
               fig.width = 12
               #fig.show = 'hide'
               )
invisible(purl("Marwick_Hayes_et_al.Rmd",  
               output="temp",
               quiet=TRUE))
read_chunk("temp")
```


```{r, ref.label="load_libraries"}
```

```{r, ref.label="load_data"}
```

```{r, ref.label="split_by_trampling_event", fig.show = 'hide'}
```

```{r, ref.label="prepare_point_data", fig.show = 'hide'}
```

```{r, ref.label="zinggplotofartefacts", fig.show = 'hide'}
```

```{r, ref.label = "hordispldists", fig.show = 'hide'} 
```

```{r, ref.label = "hordirectionshist", fig.show = 'hide'}
```

```{r, ref.label = "hormodels", fig.show = 'hide'}
```

## Visualising individual artefact movements

In the plot below we can see the vertical distance each artefact has moved after each trampling event.  


```{r, ref.label = "verdisp", fig.show = 'hide'}
```


```{r}
artefact_num <- 33
ver_indiv_plot <- 
ver %>% 
  gather(variable, 
         value,
         -artefact) %>% 
  mutate(variable = ifelse(variable == 'five', 5,
                        ifelse(variable == 'ten', 10,
                          ifelse(variable == 'fifteen', 15, NA))))

ggplot(ver_indiv_plot, aes(variable, 
             value,
             group = artefact)) +
  geom_point() +
  geom_text_repel(data = ver_indiv_plot[ver_indiv_plot$variable == 10,],
                  aes(label = artefact)) +
  geom_line() +
  ylab("Vertical displacement") +
  xlab("Trampling event (mins)") +
  scale_x_continuous(breaks = c(5, 10, 15),
                     labels = c(5, 10, 15)) +
  theme_minimal()




```



## Diagnostic plots for linear models of horizontal displacement of artefacts

```{r, ref.label = "hormodelsdiagnostics", fig.show = 'hide'}
```

```{r,  fig.show = 'as.is'}
# draw diagnostic plots
hormodels_diagnostics_plots <- 
 map(hormodels_diagnostics,
     ~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

# view
hormodels_diagnostics_plots


kable(hormodels_diagnostics_influence, 
      caption = "Horizontal displacement: Correlation of artefact volume with Cook's distance values for each model")
  
```

## Diagnostic plots for linear models of horizontal direction of artefacts

```{r, ref.label = "hordirmodelsdiagnostics", fig.show = 'hide'}
```

```{r,  fig.show = 'as.is'}

 # draw diagnostic plots
  
hordirmodels_diagnostics_plots <- 
 map(hordirmodels_diagnostics, ~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

# view
hordirmodels_diagnostics_plots



kable(hordirmodels_diagnostics_influence, 
      caption = "Horizontal direction: Correlation of artefact volume with Cook's distance values for each model")
 
```

## Diagnostic plots for linear models of vertical displacement of artefacts

```{r, ref.label = "vermodels", fig.show = 'hide'}
```

```{r, ref.label = "vermodelsdiagnostics", fig.show = 'hide'}
```

```{r,  fig.show = 'as.is'}

# draw diagnostic plots
vermodels_diagnostics_plots <- 
  map(vermodels_diagnostics, 
      ~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

vermodels_diagnostics_plots


kable(vermodels_diagnostics_influence, 
      caption = "Vertical displacement: Correlation of artefact volume with Cook's distance values for each model")
```

## Diagnostic plots for linear models of artefact orientation

```{r, ref.label = "verangles", fig.show = 'hide'}
```

```{r, ref.label = "artefactorientation", fig.show = 'hide'}
```

```{r, ref.label = "artefactplunge", fig.show = 'hide'}
```

```{r, ref.label = "orientplungcorr", fig.show = 'hide'}
```

```{r, ref.label = "orientmodelsdiagnostics", fig.show = 'hide'}
```

```{r,  fig.show = 'as.is'}
orientmodels_diagnostics_plots <- 
  orientmodels_diagnostics %>% 
  map(~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

orientmodels_diagnostics_plots


kable(orientmodels_diagnostics_influence, 
      caption = "Artefact orientation: Correlation of artefact volume with Cook's distance values for each model")
 
```

## Diagnostic plots for linear models of artefact plunge


```{r, ref.label = "plungemodelsdiagnostics", fig.show = 'hide'}
```


```{r,  fig.show = 'as.is'}

plungemodels_diagnostics_plots <- 
  plungemodels_diagnostics %>% 
  map(~autoplot(., 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10))

plungemodels_diagnostics_plots


kable(plungemodels_diagnostics_influence, 
      caption = "Artefact plunge: Correlation of artefact volume with Cook's distance values for each model")
 
```


## Power analysis for the linear models



```{r}
library(pwr)
library(purrr)
library(dplyr)
pwr_test <- 
pwr.f2.test( 
  u = 1,     # n-1 for 1 independent variable plus the intercept
  v = 29,    # n-p 30-1
  f2 = 0.35, # Cohen suggests f2 values of 0.02, 0.15, and 0.35 represent small, medium, and large effect sizes.
  sig.level = 0.05,
  power = NULL)
```

With thirty artefacts, the linear models in this paper have a power value of `r round(pwr_test$power, 3)`. This is a high power value, and indicates that our sample size is adequete for our regression models. 

## Simulation of many trampling events

How to determine the how deep is too deep for trampling to have an effect on an artefact? Here we look at the distribution of distances below the starting point, and assess which theoretical distribution is the best fit. 

```{r}
ggplot(ver.s, aes(displ)) + 
  geom_histogram()

# take the -ve values and fit a distribution
# focus on the distrubution of moves when an artefact moves down
negative_vert <- ver.s$displ[ver.s$displ < 0]

library(fitdistrplus)

descdist(negative_vert, discrete = FALSE)

possible_distributions <- c("lnorm", "gamma",  "beta", "exp", "norm")

# compute fits to distributions
fits <- map(possible_distributions, ~fitdist(-negative_vert, .x))

# for visual inspections
map(fits, ~plot(.x))

# find model with lowest AIC
best_distribution <- 
map_dbl(fits, ~.x$aic) %>% 
  which.min() %>% 
  possible_distributions[.]

best_distribution_details <- 
fits[[which(possible_distributions == best_distribution)]]
```

We evaluated `r length(possible_distributions)` possible distributions to find the best fit with our observed values of artefact displacement below the point of origin. Using visual inspection of diagnostic plots and the AIC valuess, we have determined that a `r best_distribution` distribution is the best fit for the observed vertical displacenent values.

```{r ref.label = "sim", fig.show = 'hide'}
```

```{r}
# generate a bunch of fitted gamma distributions
n <- 1000
many_fitted_distributions <- 
replicate(n, rgamma(n = length(negative_vert), 
                      shape = best_distribution_details$estimate[[1]], 
                      rate= best_distribution_details$estimate[[2]]))



data_to_plot <- 
   data.frame(many_fitted_distributions,
            observed_data = -negative_vert) %>% 
   gather(variable, 
          value)

ggplot() +
   geom_density(data = filter(data_to_plot, 
                              grepl("X", variable)),
                aes(value, 
                    group = variable),
               colour = alpha("grey70", 0.1)) +
  geom_density(data = filter(data_to_plot, 
                             variable == 'observed_data'),
               aes(value),
               colour = alpha("red", 1)) +
   theme_minimal() +
   xlab("Depth below starting location (m)")
```

The plot above shows `r n` simulated `r best_distribution` distributions (in grey) and our observed distrubtion in red. 

The plot below shows the movement of one artefact during `r n` simulated trampling events. 

```{r}
ninety_nine_point_five <- 
qgamma(0.995, 
       shape = best_distribution_details$estimate[[1]],
       rate = best_distribution_details$estimate[[2]])
```


```{r fitteddistsom, ref.label='fitteddist'}
```


```{r simplottingoneartefactsom}
# move one artefact n times and look at the results

n <- 1000
artefact_num = 31
move_artefact_n <- 
  move_the_artefact(artefact_num, 
                    artefact_event_coords_diffs,
                    n = n)

# get some variables to set the axis limits of the plot

# range of hor movement
rnge_n <- max(move_artefact_n$Northing) - min(move_artefact_n$Northing)
rnge_e <- max(move_artefact_n$Easting) - min(move_artefact_n$Easting)
rngs <- c(rnge_n, rnge_e)
max_extent <- round(rngs[which.max(rngs)], 2)

# hor xmin, ymin, etc. 
const <- 1.2
xmin <- median(move_artefact_n$Northing) - max_extent/const
xmax <- median(move_artefact_n$Northing) + max_extent/const
ymin <- median(move_artefact_n$Easting) - max_extent/const
ymax <- median(move_artefact_n$Easting) + max_extent/const

# range of vert movement
rnge_vert <- max(move_artefact_n$Elevation) - min(move_artefact_n$Elevation)
rngs_v <- c(rnge_n, rnge_vert)
max_extent_v <- round(rngs_v[which.max(rngs_v)], 2)

# vert xmin, ymin, etc. 
vconst <- 0.5
vxmin <- median(move_artefact_n$Northing) - max_extent_v/vconst
vxmax <- median(move_artefact_n$Northing) + max_extent_v/vconst
vymin <- median(move_artefact_n$Elevation) - max_extent_v/vconst
vymax <- median(move_artefact_n$Elevation) + max_extent_v/vconst

# plot plan to see horizontal movement
hplot <- 
ggplot(move_artefact_n,
       aes(Northing,
           Easting)) +
  geom_point(alpha = 0.1) +
  geom_point(data = move_artefact_n[1,],
       aes(Northing,
           Easting),
       colour = "red",
       size = 3)  +
 coord_fixed(ratio = 1) +
  xlim(xmin, xmax) +
  ylim(ymin, ymax) +
  ggtitle(paste0("Plan view of all artefact locations \nduring ", n, " simulated events for artefact ", artefact_num))

# plot section to see vertical movement
vplot <- 
ggplot(move_artefact_n,
       aes(Northing,
           Elevation)) +
  geom_point(data = move_artefact_n[1,],
       aes(Northing,
           Elevation),
       colour = "red",
       size = 3)  +
  geom_point(alpha = 0.1) +
 coord_fixed(ratio = 1) +
  xlim(vxmin, vxmax) +
  ylim(vymin, vymax) +
  ggtitle(paste0("Section view of all artefact locations \nduring ", n, " simulated events for artefact ", artefact_num))

grid.arrange(hplot, vplot, ncol =  2)
```


```{r, eval = FALSE}
# run the function to move all the artefacts, and keep the results after each event so we can see the cumulative effect. 

n <- 10 # how many times to move each artefact
move_keep_each <- vector("list", n)
for(i in 1:n){
  move_keep_each[[i]] <- 
move_all_the_artefacts(n = i, artefact_event_coords_diffs)
}

move_keep_each <- 
lapply(move_keep_each, function(i) i[i$event == max(i$event),])

move_keep_each_hdist <-
  lapply(move_keep_each, function(i) 
  sqrt(
    (artefact_event_coords_diffs$Northing.1.B - i$Northing)^2 +
    (artefact_event_coords_diffs$Easting.1.B  - i$Easting)^2
 ))

move_keep_each_hdist_df <- data.frame(do.call(cbind, move_keep_each_hdist))
move_keep_each_hdist_df_stack <- stack(move_keep_each_hdist_df)
move_keep_each_hdist_df_stack$event <- 
  as.numeric(move_keep_each_hdist_df_stack$ind)
move_keep_each_hdist_df_stack$event <-
  factor(move_keep_each_hdist_df_stack$event,
         levels = 1:n)

# plot h dist cumulative
hdist_cum_plot <- 
ggplot(move_keep_each_hdist_df_stack,
       aes(values,
           group = event)) +
  geom_density( alpha = 0.05, linetype = 0, fill = "grey20") +
  xlab(paste0("Horizontal distance (m) from starting point at the end of each of ", n, " events"))

# same for vertical distances
move_keep_each_vdist <-
  lapply(move_keep_each, function(i) 
  artefact_event_coords_diffs$Elevation.1.B - i$Elevation)

move_keep_each_vdist_df <- data.frame(do.call(cbind, move_keep_each_vdist))
move_keep_each_vdist_df_stack <- stack(move_keep_each_vdist_df)
move_keep_each_vdist_df_stack$event <- 
  as.numeric(move_keep_each_vdist_df_stack$ind)
move_keep_each_vdist_df_stack$event <-
  factor(move_keep_each_vdist_df_stack$event,
         levels = 1:n)

# plot v dist cumulative
vdist_cum_plot <- 
ggplot(move_keep_each_vdist_df_stack,
       aes(values,
           group = event)) +
  geom_density( alpha = 0.02, linetype = 0, fill = "grey20") +
  xlab(paste0("Vertical distance (m) from starting point at the end of each of ", n, " events"))

grid.arrange(hdist_cum_plot, vdist_cum_plot, ncol = 1)
```




```{r, eval=FALSE}
## Explore Bayesian linear models

# summarise multiple models hor dir
library(rstanarm)

tidy_stan_lm <- function(x){
  df <- data.frame(estimate = x$coefficients[-1],
        t(quantile(rstanarm:::as.matrix.stanreg(x)[,2], 
                              probs = c(0.05, 0.95))), 
        std.error = x$ses[-1], 
        R2 = mean(x$stanfit@sim$samples[[1]]$`R2[1]`))
  names(df) <- c("estimate", "estimate_5", "estimate_95", "std.error", "R2")
  df
}


# the broom::tidy method isn't very useful 
tidy_stan_lm <- function(x){
  df <- data.frame(estimate = x$coefficients[-1],
        t(quantile(rstanarm:::as.matrix.stanreg(x)[,2], 
                              probs = c(0.05, 0.95))), 
        std.error = x$ses[-1], 
        R2 = mean(x$stanfit@sim$samples[[1]]$`R2[1]`))
  names(df) <- c("estimate", "estimate_5", "estimate_95", "std.error", "R2")
  df
}


stan_linear_models_estimates_hor_dir <- 
hor_dir_artefact %>% 
  gather(value, 
         variable, 
         -angle_hor_displacement) %>% 
  filter(value %in% vars_of_interest) %>% 
  group_by(value) %>% 
  do(tidy_stan_lm(stan_lm(scale(angle_hor_displacement) ~ scale(variable), 
                    data = .,
                    prior = NULL,
                    chains = 1, 
                    cores = parallel::detectCores(), 
                    seed = 1))) 


stan_linear_models_estimates_hor_dir <- 
stan_linear_models_estimates_hor_dir %>% 
  ungroup %>% 
  mutate(value = gsub("\\(mm\\)", "", value),
         model = "Horizontal direction")

hor_disp_artefact <- merge(hor, tramp, by = "artefact")

# reshape to long, keeping the 15 mist hor dist out

library(tidyr)
hor_disp_artefact_long <- gather(hor_disp_artefact, 
                                 value, variable, 
                                 -fifteen)


# summarise multiple models hor dipls
stan_linear_models_estimates_hor_disp <- 
hor_disp_artefact_long %>% 
  filter(value %in% vars_of_interest) %>% 
  group_by(value) %>% 
  do(tidy_stan_lm(stan_lm(scale(fifteen) ~ scale(variable), 
                    data = .,
                    prior = NULL,
                    chains = 1, 
                    cores = parallel::detectCores(), 
                    seed = 1))) %>% 
  ungroup %>% 
  mutate(value = gsub("\\(mm\\)", "", value),
         model = "Horizontal distance")

# combine two models for hor dir and dist to plot
stan_hor_models <- 
rbind(stan_linear_models_estimates_hor_dir,
      stan_linear_models_estimates_hor_disp)

# estimate distribution includes zero?
stan_hor_models <- 
  stan_hor_models %>% 
  mutate(zero = ifelse(0 < estimate_95 & estimate_5 < 0, "yes", "no"))
   
ggplot() +
    geom_point(data = stan_hor_models,
        aes(estimate, 
            R2,
            colour = zero)) +
    geom_vline(xintercept = 0) +
    geom_errorbarh(data = stan_hor_models, 
                  aes(estimate, 
                      R2,
            xmin = estimate_5, 
            xmax = estimate_95, 
            height = 0)) +
    geom_text_repel(data = stan_hor_models,
                    aes(estimate, 
                    R2,
                    label = value),
            nudge_x = 0.02,
            nudge_y = 0.02) +
   theme_bw() +
   xlab("Standardised estimates with 90% confidence intervals") +
   ylab(expression(R^2)) +
   scale_size("p-values", 
              breaks = c(1/0.75, 1/0.25, 1/0.05, 1/0.01, 1/0.005), 
              labels = c(0.75, 0.25, 0.05, 0.01, 0.005)) +
  guides(size = guide_legend(override.aes = list(pch = c(19, 19, 21, 21, 21)))) +
  facet_wrap(~model, ncol = 2)

```



```{r, eval=FALSE}
library(rstanarm)
# the broom::tidy method isn't very useful 
tidy_rstan_lm <- function(model){
  # compute loo: Leave-One-Out (LOO) cross-validation
  loo_model <- loo::loo(model)
  # tidy the output
  df <- data.frame(
          estimate = model$coefficients[-1],
          t(quantile(rstanarm:::as.matrix.stanreg(model)[,2], 
                                probs = c(0.05, 0.95))), 
          std.error = model$ses[-1], 
          R2 = mean(model$stanfit@sim$samples[[1]]$`R2[1]`),
          looic = unname(loo_model$looic))
  names(df) <- c("estimate", "estimate_5", "estimate_95", "std.error", "R2", "looic")
  df
}

augment_rstan_lm <- function(model){
  # compute the model
  # compute loo: Leave-One-Out (LOO) cross-validation
  loo_model <- loo::loo(model)
  df <- data.frame(loo_model$pointwise)
  df
}

# total vertical displacement, don't care if up or down
stan_linear_models_estimates_unsigned <- 
ver_disp_artefact_long %>% 
  filter(value %in% vars_of_interest) %>% 
  nest(-value) %>% 
  mutate(model = map(data, ~stan_lm(abs(fifteen) ~ variable, data = .,
                    prior = NULL, 
                    chains = 1, 
                    cores = 2, 
                    seed = 1))) 

stan_linear_models_estimates_unsigned %>% 
  mutate(augmented = map(.$model, ~augment_rstan_lm(.)))


%>% 
  ungroup %>% 
  mutate(value = gsub("\\(mm\\)", "", value)) %>% 
  mutate(signed = "unsigned")

#  vertical displacement relative to zero, do care if up or down
stan_linear_models_estimates_signed <- 
ver_disp_artefact_long %>% 
  filter(value %in% vars_of_interest) %>% 
  group_by(value) %>% 
  do(tidy_stan_lm(stan_lm(scale(fifteen) ~ scale(variable), 
                    data = .,
                    prior = NULL,
                    chains = 1, 
                    cores = parallel::detectCores(), 
                    seed = 1))) %>% 
  ungroup %>% 
  mutate(value = gsub("\\(mm\\)", "", value)) %>% 
  mutate(signed = "signed")

stan_all_models_ver <- 
rbind(stan_linear_models_estimates_unsigned,
      stan_linear_models_estimates_signed)

ggplot(stan_all_models_ver, 
        aes(estimate, 
            R2, 
            height = 0)) +
    geom_point(data = stan_all_models_ver,
        aes(estimate, 
            R2)) +
    geom_vline(xintercept = 0) +
    geom_errorbarh(data = stan_all_models_ver, 
                  aes(estimate, 
                      R2,
            xmin = estimate_5, 
            xmax = estimate_95, 
            height = 0)) +
    geom_text_repel(aes(estimate, 
            R2,
            label = value),
            nudge_x = 0.02,
            nudge_y = 0.02) +
   theme_bw() +
   xlab("Standardised estimates with 90% confidence intervals") +
   ylab(expression(Adjusted~R^2)) +
   scale_size("p-values", 
              breaks = c(1/0.75, 1/0.25, 1/0.05, 1/0.01, 1/0.005), 
              labels = c(0.75, 0.25, 0.05, 0.01, 0.005)) +
  guides(size = guide_legend(override.aes = list(pch = c(19, 19, 21, 21, 21)))) +
  facet_wrap(~ signed, ncol = 2)

```


```{r}
# delete the temp file we made at the top of this file
unlink("temp")
```

# Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon2, cache = FALSE}
# which R packages and versions?
devtools::session_info()

# what commit is this file at?
library(git2r)
repo <- repository(path = "../..")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is "`r last_commit@summary`". 
